Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-07 main.tex
passe section finale parsing

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$, 
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA 
$B_{T, A}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$, 
$B_{T, A}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init_T, \bar{\wei}, \final_T >$,
where $\bar{\wei}$ contains $\wei_{10}$, $\wei_{01}$, and $\wei_{11}$,
from $Q \times Q$ into respectively 
$\Phi_{\Sigma}$, $\Phi_{\Delta}$, and $\Phi_{\Sigma, \Delta}$,
and let $A = \< P, \init_A, \wei_1, \final_A >$
with $\wei_1: Q \times Q \to \Phi_{\Sigma}$.
%
\noindent
The state set of $B_{T, A}$ will be $Q' = P \times Q$.
The entering, leaving and transition functions of $B_{T, A}$ will 
simulate synchronized computations of $A$ and $T$, 
while reading an output word of $\Delta^*$.
%
Its state entering functions is defined 
for all $p \in P$, $q \in Q$ 
by $\init'(p, q) = \init_A(p) \otimes \init_T(q)$.
The transition function $\wei'_1$ will roughly perform 
a synchronized product of transitions defined by $\wei_1$, 
$\wei_{01}$ ($T$ reading in output word and not an input word)
and $\wei_{11}$ ($T$ reading both an input word and an output word).
%
Moreover, $\wei'_1$ also needs to simulate transitions 
defined by $\wei_{10}$: $T$ reading in input word and not an output word.
Since $B_{T, A}$  will read only in the output word, such a transition corresponds
to an $\varepsilon$-transition of $\SWA$, 
but $\SWA$ have been defined without $\varepsilon$-transitions.
Therefore, in order to take care of this case, we perform an on-the-fly
suppression of $\varepsilon$-transition in the $\SWA$ in construction, 
following the algorithm of~\cite{LombardySakarovitch12ciaa}. 
%
% Initialize state leaving functions i
%and $\final'(p, q) = \final_A(p) \otimes \final_T(q)$.

%Every transition of $B_{T, A}$ will
%simulate a sequence of transitions of $T$ performing the following steps:
%advance in the input word while staying immobile in the output word, 
%and then make one step in the output word (and advance in the input word or not).

@@

%The complexity of construction
\noindent
The construction time and size for $B_{T, A}$ are $O(\| T \|^3 . \| A \|^2)$,
where the sizes $\| T \|$ and $\| A \|$ are their number of states.
%of~$T$ is its number of states $|Q|$.
\florent{revise with nb of tr. and states}

\begin{corollary} \label{cor:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$, 
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and $s \in \Sigma^+$, 
there exists an effectively constructible \SWA 
$B_{T, s}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$, $B_{T, s}(t) = T(s, t)$.
\end{corollary}

@@

Let us now apply the models and results of the previous sections %in order to define 
to the problem of parsing over infinite alphabet. %appropriate
%
%\subsection{Definition}
%
Let~$\Sigma$ be a countable input alphabet, 
and~$\Omega = \Omegai \uplus \Omegac \uplus \Omegar$ be a countable output alphabet.
%
Let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a 
commutative, bounded, and complete \florent{total?} semiring  
and let $\bar\Phi$ be an effective label theory over $\Semiring$,
containing $\Phi_\Sigma$, $\Phi_{\Sigma, \Omegai}$, as well as
$\Phii$, $\Phic$, $\Phir$, $\Phicr$
(following the notations of Section~\ref{sec:SWVPA-def}).
%
\noindent
We assume given the following input:
\begin{description}
\item[--] a \SWT $T$ over $\Sigma$, $\Omegai$, $\Semiring$, and $\bar\Phi$, 
defining a measure %between words 
$T: \Sigma^* \times \Omegai^* \to \Semiring$,

@@

For all $s \in \Sigma^*$ and $t \in \Omega^*$, let 
\(d(s, t) = T\bigl(s, t|_\Omegai \bigr)\), 

@@

Every labeled tree $t$ can be linearized into a nested word $\lin(t) \in \Omega^*$, 
assuming \eg that $\Omegai$ contain the symbols labelling the leaves (symbols of rank $0$) 

@@

$\ccall{b}$ and $\creturn{b}$ for each symbol $b$ labelling inner nodes (symbols of rank $>0$). 

@@

sequence of leaves of $t$, enumerated in a \emph{dfs}-traversal. 

@@

into a $\SWVPA$ computing $A(\lin(t))$ for every tree $t$. 
That also holds, for the set of ASTs of a weighted CF-grammar.

@@

Therefore, $\SW$-parsing generalizes 
the problem of searching, the best derivation of a weighted CF-grammar  
that yields a given input, sometimes referred as~\emph{weighted parsing},
%corresponds to $\SW$-parsing in the case of finite alphabets
%and when the transducer $T$ characterizes identity
see \eg~\cite{Goodman99SemiringParsing} 
and~\cite{MorbitzVogler19weighted-parsing} for a more general weighted parsing framework.
%weighted parsing and convention parsing 
The latter indeed corresponds to the 
particular case where the alphabet is finite, 
$T$ computes identity and $A$ is obtained from the weighted CF grammar.

%\subsection{Computation}
%
\begin{proposition}
The problem of Symbolic Weighted  parsing 
can be solved in PTIME in the size of the input \SWT $T$, \SWVPA $A$
and input word $s$, 
and the computation time of the functions of the label theory.
\end{proposition}
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, also called parsing by intersection.

\noindent
We first extend the \SWT $T$ over $\Sigma$, $\Omegai$, $\Semiring$, and $\bar\Phi$, 
into a \SWT $T'$ over $\Sigma$ and $\Omega$ (and the same semiring and label theory),
such that for all $s \in \Sigma^*$, and $u \in {\Omega}^*$, 
$T'(s, u) = T(s, u|_{\Omegai})$. The transducer $T'$ simply skips every symbol 
$b \in {\Omega} \setminus \Omegai$, 
by the addition to the transition of $T$,
of new transitions of the form $\wei_{01}(q, \varepsilon, b, q')$.
%
\noindent
Then, given an input word $s \in \Sigma^*$, 
using Corolary~\ref{cor:epsilon}, 
we compute the \SWA $B_{T', s}$, 
such that for all $t \in \Omega^*$, $B_{T', s}(t) = d(s, t)$.
%
%This automaton is such that for all $t \in \T(\Omega)$, 
%\[ 
%   A_{T', s}\bigl(\lin(t)\bigr) 
% = T'\bigl(s, \lin(t)\bigr) 
% = T'\bigl(s, \lin(t)|_{\Omegai}\bigr) 
% = d(s, t).
%\]

\noindent
Next, %we convert the input \SWTA $A$ over $\Omega$
%into a \SWVPA $A'$ over $\hat\Omega$, using Proposition~\ref{lem:SWTA}, and 
we compute the \SWVPA $B_{T', s} \otimes A$, 
using Proposition~\ref{prop:SWVPA-product}.
%
\noindent
It remains to compute a best nested-word $w \in {\Omega}^*$ 
using the best-search procedure of Proposition~\ref{th:best-search}.
%and convert it into a best tree in $\T(\Omega)$ in order to solve SW parsing
%for~$T$, $A$ and~$s$.
%\qed  
\end{proof}

>>>>

\begin{proposition} \label{prop:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$, 
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and a $\SWA$ $A$ over $\Sigma$, $\Semiring$ and $\bar\Phi$,
there exists an effectively constructible \SWA 
$B_{A, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$, 
$B_{A, T}(t) = \displaystyle\bigoplus_{s\in \Sigma^*} A(s) \otimes T(s, t)$.
\end{proposition}
%
\begin{proof}
Let $T = \< Q, \init_T, \bar{\wei}, \final_T >$,
where $\bar{\wei}$ contains $\wei_{10}$, $\wei_{01}$, and $\wei_{11}$,
from $Q \times Q$ into respectively 
$\Phi_{\Sigma}$, $\Phi_{\Delta}$, and $\Phi_{\Sigma, \Delta}$,
and let $A = \< P, \init_A, \wei_1, \final_A >$
with $\wei_1: Q \times Q \to \Phi_{\Sigma}$.
%
\noindent
The state set of $B_{A, T}$ will be $Q' = P \times Q$.
The entering, leaving and transition functions of $B_{A, T}$ will 
simulate synchronized computations of $A$ and $T$, 
while reading an output word of $\Delta^*$.
%
Its state entering functions is defined 
for all $p \in P$, $q \in Q$ 
by $\init'(p, q) = \init_A(p) \otimes \init_T(q)$.
The transition function $\wei'_1$ will roughly perform 
a synchronized product of transitions defined by $\wei_1$, 
$\wei_{01}$ ($T$ reading in output word and not an input word)
and $\wei_{11}$ ($T$ reading both an input word and an output word).
%
Moreover, $\wei'_1$ also needs to simulate transitions 
defined by $\wei_{10}$: $T$ reading in input word and not an output word.
Since $B_{A, T}$  will read only in the output word, such a transition corresponds
to an $\varepsilon$-transition of $\SWA$, 
but $\SWA$ have been defined without $\varepsilon$-transitions.
Therefore, in order to take care of this case, we perform an on-the-fly
suppression of $\varepsilon$-transition in the $\SWA$ in construction, 
following the algorithm of~\cite{LombardySakarovitch12ciaa}. 
%
% Initialize state leaving functions i
%and $\final'(p, q) = \final_A(p) \otimes \final_T(q)$.

%Every transition of $B_{A, T}$ will
%simulate a sequence of transitions of $T$ performing the following steps:
%advance in the input word while staying immobile in the output word, 
%and then make one step in the output word (and advance in the input word or not).

@@

%The complexity of construction
\noindent
The construction time and size for $B_{A, T}$ are $O(\| T \|^3 . \| A \|^2)$,
where the sizes $\| T \|$ and $\| A \|$ are their number of states.
%of~$T$ is its number of states $|Q|$.
\florent{revise with nb of tr. and states}

\begin{corollary} \label{cor:epsilon}
Given a \SWT $T$ over $\Sigma$, $\Delta$, 
$\Semiring$ commutative, bounded and complete,
and $\bar\Phi$ effective,
and $s \in \Sigma^+$, 
there exists an effectively constructible \SWA 
$B_{s, T}$ over $\Delta$, $\Semiring$ and $\bar\Phi$,
such that for all $t \in \Delta^*$, $B_{s, T}(t) = T(s, t)$.
\end{corollary}

@@

Let us now apply the models and results of the previous sections %in order to define 
to the problem of parsing over an infinite alphabet. %appropriate
%
%\subsection{Definition}
%
Let~$\Sigma$ 
and~$\Omega = \Omegai \uplus \Omegac \uplus \Omegar$ 
be countable input and output alphabets,
let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a 
commutative, bounded, and complete \florent{total?} semiring  
and let $\bar\Phi$ be an effective label theory over $\Semiring$,
containing $\Phi_\Sigma$, $\Phi_{\Sigma, \Omegai}$, as well as
$\Phii$, $\Phic$, $\Phir$, $\Phicr$
(following the notations of Section~\ref{sec:SWVPA-def}).
%
\noindent
We assume given the following input:
\begin{description}
\item[--] a \SWT $T$ over $\Sigma$, $\Omegai$, $\Semiring$, and $\bar\Phi$, 
defining a measure %between words 
$T: \Sigma^* \times \Omegai^* \to \Semiring$,

@@

For all $u \in \Sigma^*$ and $t \in \Omega^*$, let 
\(d(u, t) = T\bigl(u, t|_\Omegai \bigr)\), 

@@

%\subsection{Computation}
\begin{proposition}
The problem of Symbolic Weighted  parsing 
can be solved in PTIME in the size of the input \SWT $T$, \SWVPA $A$
and input word $s$, 
and the computation time of the functions and operators of the label theory.
\end{proposition}
\begin{proof} (sketch)
We follow a \emph{Bar-Hillel} construction, for parsing by intersection.
Let us first extend the \SWT $T$ over $\Sigma$, $\Omegai$ 
into a \SWT $T'$ over $\Sigma$ and $\Omega$ 
(and the same semiring and label theory $\Semiring$ and $\bar\Phi$),
such that for all $u \in \Sigma^*$, and $t \in {\Omega}^*$, 
$T'(u, u) = T(u, t|_{\Omegai})$. 
The transducer $T'$ simply skips every symbol 
$b \in {\Omega} \setminus \Omegai$, 
by the addition to $T$,
of new transitions of the form $\wei_{01}(q, \varepsilon, b, q')$.
Then, using Corolary~\ref{cor:epsilon}, 
we construct from the input word $s \in \Sigma^*$ and $T'$
a \SWA $B_{s, T'}$, 
such that for all $t \in \Omega^*$, $B_{s, T'}(t) = d(s, t)$.
%This automaton is such that for all $t \in \T(\Omega)$, 
%\[ 
%   A_{T', s}\bigl(\lin(t)\bigr) 
% = T'\bigl(s, \lin(t)\bigr) 
% = T'\bigl(s, \lin(t)|_{\Omegai}\bigr) 
% = d(s, t).
%\]
Next, %we convert the input \SWTA $A$ over $\Omega$
%into a \SWVPA $A'$ over $\hat\Omega$, using Proposition~\ref{lem:SWTA}, and 
we compute the \SWVPA $B_{s, T'} \otimes A$, 
using Proposition~\ref{prop:SWVPA-product}.
It remains to compute a best nested-word $t \in {\Omega}^*$ 
using the best-search procedure of Proposition~\ref{th:best-search}.
%and convert it into a best tree in $\T(\Omega)$ in order to solve SW parsing
%for~$T$, $A$ and~$s$.
%\qed  
\end{proof}
The $\SW$-parsing generalizes 
the problem of searching the best derivation (AST) of a weighted CF-grammar  
that yields a given input word.
The latter problem, sometimes called~\emph{weighted parsing},
(see \eg~\cite{Goodman99SemiringParsing} 
 and~\cite{MorbitzVogler19weighted-parsing} for general weighted parsing frameworks)
corresponds to $\SW$-parsing in the case of finite alphabets,
a transducer $T$ computing the identity and some \SWVPA~$A$ 
obtained from the weighted CF grammar.
Indeed, the \emph{depth-first} traversal of an AST $\tau$ 
yields a well-parenthesised word $\lin(\tau)$ over an alphabet 
$\Omega = \Omegai \uplus \Omegac \uplus \Omegar$,  
assuming \eg that $\Omegai$ contain the symbols labelling the leaves of $\tau$ (symbols of rank $0$) 

@@

$\ccall{b}$ and $\creturn{b}$ for each symbol $b$ labelling inner nodes of $\tau$ (symbols of rank $>0$). 

@@

sequence of leaves of $\tau$. %, enumerated in a \emph{dfs}-traversal. 

@@

%(in particular a TA computing on the AST of a given CF-grammar)
into a $\SWVPA$ computing $A(\lin(\tau))$ for every tree $\tau$. 
That also holds for the set of ASTs of a weighted CF-grammar.

@@
